---
title: "Updating prokaryote trees"
author: "Laura Tibbs-Cortes"
date: "2024-12-16"
output: html_document
---

## Adding new prokaryote trees

As of December 2024, there are issues with the hard-coded `use_recommended_prokaryotes` function having out-of-date UniProt ids and resulting in 30 out of 85 prokaryotes having warnings. (see https://github.com/arendsee/phylostratr/issues/34) 

This markdown shows an example of using my new functions `generate_prokaryote_tree` and `use_custom_prokaryote_tree`, as well as an updated `uniprot_strata`, to make a new prokaryote tree with custom size and weights.

```{r echo=T, results='hide'}
# load libraries
library(tidyverse)
library(data.table)
library(ape) # this library is used within phylostratr

# library(devtools)
# install_github('ltibbs/phylostratr', force=T)
library(phylostratr)

source("R/update_prokaryotes.R")
```

Set focal species and weights for tree:
You will likely want to set some weights based on proteome quality to improve results.
```{r}
# Set focal species (eukaryote)
focal_taxid <- '4577' # zea mays focal species example

# set weights
# initial weights: reference proteomes are 1.1
# CAREFUL: uniprot_weight_by_ref apparently requires clade to be provided
# as a numeral rather than as a character like in other functions
my.weights <- c(uniprot_weight_by_ref(clade=2), #bacteria
                uniprot_weight_by_ref(clade=2157), # archaea
                uniprot_weight_by_ref(clade=2759)) # eukarya
my.weights['3702'] <- 1.5 # example custom weighting, increasing weight for Arabidopsis
my.weights['83333'] <- 1.5 # example custom weighting, increasing weight for E coli
```

Gnerate diverse subtrees of bacteria and archaea based on current UniProt with custom weights and desired numbers of species. This makes use of the `diverse_subtree` function of phylostratr:
```{r}
bac.tree <- generate_prokaryote_tree(weights=my.weights, 
                                     current.domain="2", #2=UniProt taxonomy id for bacteria
                                     domain.n=70) # ~70 bacterial species used in original paper
arc.tree <- generate_prokaryote_tree(weights=my.weights, 
                                     current.domain="2157", #2157=UniProt taxonomy id for archaea
                                     domain.n=15) # ~15 archaeal species used in original paper

# combine the prokaryotic trees 
# keep them in phylo format because Strata format 
# had issues when joining more than 1 branch onto a tree
pro.tree=bind.tree(bac.tree, arc.tree)  
```

Now, build the full strata tree by adding the prokaryotes to the focused eukaryote tree:
```{r}
# Make focused tree around taxon of interest
strata.cellular.organisms <- uniprot_strata(focal_taxid,  # get stratified relatives in UniProt
                                            from=1) %>% # stratum to begin from, where 1 is 'cellular organisms' (2 is default). Updated function as of Dec 2024 to enable it to work with from=1
  strata_apply(f=diverse_subtree, 
               n=5, 
               weights=my.weights) %>%
  use_custom_prokaryote_tree(prokaryote.phylo=pro.tree) # add the custom prokaryote tree to the focused tree
```

Visualize resulting tree:

```{r}
plot(strata.cellular.organisms%>%
       # strata_convert(target='all', to='name') %>% 
       sort_strata)
```

