---
title: "*Arabidopsis thaliana*: proteome quality estimates"
author: "Zebulun Arendsee"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{*Arabidopsis thaliana*: proteome quality estimates}
  %\VignetteEngine{knitr::rmarkdown}
---

The purpose of this vingette is to explore the use of phylostratigraphy in
assessing proteome quality. The general idea is this: if a protein in a
well-annotated proteome has a matches against ancient strata, then it should
tend to have matches against younger strata. Of course deletions are possible
and the deep matches could have been false positives.

We will explore this idea in the context of the *Arabidopsis thaliana*. We will
reuse old results. 

```{r, eval=FALSE}
data(arabidopsis_strata)
data(arabidopsis)
strata <- arabidopsis_strata
hittable <- arabidopsis
```

We will need a distance tree. But the NCBI common tree does not have this
information. The responsible thing to do would be to build a new tree from our
sequence or search for a high quality one in literature. But we will instead 

```{r, eval=FALSE}
library(taxize)
library(taxizedb)
library(dplyr)

lineages <- taxizedb::classification(leafs(strata), db='ncbi')
a <- lineages_to_phylo(lineages)
b <- taxize::class2tree(lineages, check=FALSE)$phylo
b <- ape::di2multi(b)
map <- distinct(do.call(rbind, lineages)[, c('name', 'id')])
name2id <- map$id 
names(name2id) <- map$name
b$tip.label <- name2id[b$tip.label]
b <- map_node_label(a, b)
b <- root(b, node=clean_phyid(b, '131567', type='name'))
b <- ape::multi2di(b)
```

```{r, eval=FALSE}
## Uses Felsenstein's contrast approach to dependence handling, need to rethink
## whether it is applicable

# library(ape)
#
# dd <- dplyr::group_by(hittable, staxid) %>%
#     dplyr::mutate(evalue = ifelse(is.na(evalue), 10, evalue)) %>%
#     dplyr::summarize(hits = sum(evalue < 1e-5))
# x <- dd$hits
# names(x) <- as.character(dd$staxid)
#
# b2 <- subtree(b, '33090', type='name')
# x <- rbinom(length(leafs(b2)), 500, 0.8)
# names(x) <- leafs(b2, byname=TRUE)
#
# ape::pic(x, b2)
#
# # compare_seqid <- function(tree, d, i, under){
# #     foo <- subset(d, qseqid == i)
# #     x <- foo$score
# #     names(x) <- foo$staxid
# #     tree2 <- subtree(tree, under, type='index')
# #     x <- x[as.character(tree2$tip.label)]
# #     names(x) <- as.character(tree2$tip.label)
# #     x <- ifelse(is.na(x) | x < 100, 0, 1)
# #     ape::pic(x, tree2)
# # }
# #
# # hittable <- subset(hittable, staxid %in% b$tip.label)
# #
# # compare_seqid(
# #     tree  = b,
# #     d     = hittable,
# #     i     = 'AT1G09900.1',
# #     under = clean_phyid(b, '33090', type='name')
# # )
#
```
